---
title: "Broken-stick methods"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
link-external-newwindow: true 
vignette: >
  %\VignetteIndexEntry{Broken-stick methods}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
number-sections: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
#| message: FALSE
#| warning: FALSE
#| output: FALSE

library(morphmat)
library(chngpt)
library(segmented)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: setup-2
#| echo: false

mytheme <- theme_classic() + #define custom theme for ggplots
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
    text = element_text(size = 13))
```

<!-- Here I will write a vignette explaining all of the broken-stick regression methods. This will be useful for referring readers to more complex options available in packages like `segmented` that are not available within `morphmat`. -->

Sources (more will be added): Somerton [-@somerton1980]; Muggeo [-@muggeo2003; @muggeo2008; @muggeo2017]; Hall et al. [-@hall2006]

\begin{equation*}\log{y}=
    \begin{cases}
     \tilde{\beta_1} + \alpha_1 \log{x}             & \text{for }\log{x} \leq c \\
     \tilde{\beta_1} + c(\alpha_1-\alpha_2)+\alpha_2 \log{x}  & \text{for }\log{x} > c
    \end{cases}       
\end{equation*}

At $\log{x}=c$, the bottom equation becomes $\tilde{\beta_1}+(\log{x})(\alpha_1-\alpha_2)+\alpha_2\log{x}$, equivalent to the top equation. Note that in this formulation, there is only one intercept term $(\beta_1)$, so the immature and mature lines cannot have the same slope without being equivalent lines. In this model, the estimated value of the breakpoint parameter $c$ is taken to be SM50.

We will start by simulating a data set with a known SM50 value of 75 mm to demonstrate the use (and limitations) of broken-stick methods.
```{r}
#| label: generate-crabs

set.seed(123) # set seed for reproducibility

fc <- fake_crustaceans(
  error_scale = 17,
  slope = 9,
  L50 = 75,
  n = 800,
  allo_params = c(0.9, 0.25, 1.05, 0.2),
  x_mean = 85
)
```

```{r}
#| echo: false

ggplot() +
  geom_point(data = fc, aes(x, y), alpha = 0.4) +
  labs(x = "CW (mm)", y = "CH (mm)", ) +
  mytheme
```

## segmented package

A powerful and customizable method of implementing broken-stick regression is provided by the R package `segmented`, which has been cited by many papers using morphometric data to estimate size at maturity. Note that `segmented` includes three different methods to compute confidence intervals, the details of which are discussed in Muggeo [-@muggeo2017]. It is also possible to estimate 95% CIs via bootstrap resampling, but the computation time required is relatively high and the resulting estimates are around the same values as those from the "delta" and "gradient" methods included in the `segmented` package.

```{r}
lm_orig <- lm(y ~ x, data = fc)
lm_orig_seg <- segmented::segmented(lm_orig)
plot(lm_orig_seg)
```

The segmented package has three built-in methods to get 95% confidence intervals for the break point:

```{r}
#| echo: false
#| label: segmented-CIs

cat("Delta method:")
segmented::confint.segmented(lm_orig_seg, method = "delta")
cat("Score method:")
segmented::confint.segmented(lm_orig_seg, method = "score")
cat("Gradient method:")
segmented::confint.segmented(lm_orig_seg, method = "gradient")
```

We can also use an ANOVA test to compare the segmented model to a single linear model:
```{r}
#| label: segmented-anova

anova(lm_orig, lm_orig_seg)
```

```{r}
#| code-fold: true
#| label: fig-segmented-plot
#| fig-cap: "SM50 estimate for simulated crab data from the R package segmented"
#| echo: false

# Plot SM50 and CIs alongside the data
ggplot() +
  geom_rect(aes(xmin = confint.segmented(lm_orig_seg, method = "score")[[2]],
                xmax = confint.segmented(lm_orig_seg, method = "score")[[3]],
                ymin = -Inf, ymax = Inf, fill = "Score"), alpha = 0.5) +
  geom_rect(aes(xmin = confint.segmented(lm_orig_seg, method = "gradient")[[2]],
                xmax = confint.segmented(lm_orig_seg, method = "gradient")[[3]],
                ymin = -Inf, ymax = Inf, fill = "Gradient"), alpha = 0.9) +
  geom_rect(aes(xmin = confint.segmented(lm_orig_seg)[[2]],
                xmax = confint.segmented(lm_orig_seg)[[3]],
                ymin = -Inf, ymax = Inf, fill = "Delta"), alpha = 0.4) +
  geom_point(data = fc, aes(x, y), alpha = 0.4) +
  geom_vline(aes(xintercept = confint.segmented(lm_orig_seg)[[1]],
                 linetype = "Est. SM50"), # Central estimate line
             color = "#0f85a0", linewidth = 1) +
  geom_vline(aes(xintercept = 75, linetype = "True SM50"), # True SM50
                 color = "black", linewidth = 1) +
  labs(x = "CW (mm)",
       y = "CH (mm)",
       #title = "SM50 Estimate with 95% Confidence Intervals",
       fill = "95% CI est. method",
       linetype = NULL) +
  scale_fill_manual(values = c(
    "Score"    = "#f5db99",
    "Gradient" = "#d0e2af",
    "Delta"    = "#7bbcd5"
  )) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  theme_light() +
  ylim(0, NA) +
  mytheme +
  guides(linetype = guide_legend(override.aes = list(size = 10)))
```

## chngpt package

Another R package with the capability of segmented regression is called `chngpt`. This package differs slightly from `segmented` in the types of threshold models that are supported, the optimization algorithms used for parameter estimation, and the method(s) used to calculate confidence intervals for the breakpoint [@fong2017].

```{r}
#| label: chngpt1
#| error: false
#| collapse: true

fit_chngpt <- chngptm(
  formula.1 = y ~ 1,
  formula.2 =  ~ x,
  family = "gaussian",
  data = fc,
  type = "segmented",
  var.type = "default",
  weights = NULL
)
summary(fit_chngpt)
```

```{r}
#| label: fig-chngpt-plots
#| echo: false
#| fig-cap:
#| - "Broken-stick model from the package chngpt"
#| - "Log likelihood of the model at various breakpoint values"
#| - "Bootstrap 95% confidence intervals"
#| layout-ncol: 3

plot(fit_chngpt, 1)
plot(fit_chngpt, 2)
plot(fit_chngpt, 3)
```

As with the `segmented` package, it seems to have significantly underestimated SM50 compared to the true value of 75, which is not contained within the bootstrap confidence intervals. An alternative method for calculating confidence intervals ("model-based") produces very similar results to the default bootstrap method.

```{r}
#| label: chngpt-other-CI-methods
#| collapse: true

summary(
  chngptm(
    formula.1 = y ~ 1,
    formula.2 =  ~ x,
    family = "gaussian",
    data = fc,
    type = "segmented",
    var.type = "model", # note change from default to model
    weights = NULL
  )
)[["chngpt"]]
```

Adding a starting breakpoint value of 75 (i.e., initializing the search at the correct value) through the "chngpt.init" argument does not affect the estimate, although it does result in narrower confidence intervals:

```{r}
#| label: add-chngpt-init

summary(
  chngptm(
    formula.1 = y ~ 1,
    formula.2 =  ~ x,
    family = "gaussian",
    data = fc,
    type = "segmented",
    chngpt.init = 75,
    weights = NULL
  )
)[["chngpt"]]
```

This method is EXTREMELY sensitive to the "lb.quantile" and "ub.quantile" arguments, which refer to the lower and upper bounds of the search range for change point estimate, respectively (the function defaults are 0.05 and 0.95).  For example, if we change the lower quantile bound to 0.33 and the upper to 0.66, the model will return an SM50 value closer to the true value of 75 mm.

```{r}
#| label: chngpt-more-accurate

summary(
  chngptm(
    formula.1 = y ~ 1,
    formula.2 =  ~ x,
    family = "gaussian",
    data = fc,
    type = "segmented",
    var.type = "model",
    lb.quantile = 0.33,
    ub.quantile = 0.66,
    weights = NULL
  )
)[["chngpt"]][1]
```

The estimates and accuracy also change dramatically even with slightly different variations of data with the exact same underlying structure. Results from detailed investigations of the sensitivity of each modeling approach will be available in a forthcoming paper (Krasnow et al., *in prep*).

Although we will not test all of the options here, the `chngpt` package supports 14 different types of two-phase (one threshold) models, as well as one three-phase model: see [Fig. 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10722876/) in @son2021. Importantly, it also supports generalized linear models and linear mixed models, so you can fit models utilizing any family that can be passed to glm and include other covariates and random effects. Additional benefits of this package include several functions to simulate data and the significant capability of user-friendly customization. Options include providing your own grid of changepoints to iterate over, using five different types of bootstrap CIs with numerous options for each type, providing a bound on the slope parameters for the lines on either side of the breakpoint, and changing the search/optimization algorithm for parameter estimation to meet the needs of your data (e.g., "fastgrid2" is very fast and might be helpful for large data sets).

Using the "chngpt.test" function to run significance tests:

```{r}
#| label: chngpt-sig-tests
#| collapse: true

# Model with default upper and lower bounds
chngpt.test(
  formula.null = y ~ 1,
  formula.chngpt = ~ x,
  family = "gaussian",
  data = fc,
  type = "segmented"
)

# Model with better upper and lower bounds to return
# the true SM50 of 75 mm (0.6 and 0.3)
chngpt.test(
  formula.null = y ~ 1,
  formula.chngpt = ~ x,
  family = "gaussian",
  data = fc,
  type = "segmented",
  lb.quantile = 0.3,
  ub.quantile = 0.6
)
```

It is interesting that this function returns a slightly different breakpoint/threshold estimate from the `chngptm()` function when both are called using the default search region bounds (regardless of what set.seed is set to before calling the functions). Additionally, the test statistic for the model with the less accurate breakpoint estimate is much higher than for the more accurate model, yielding a lower p-value.

## REGRANS

We will also test the broken-stick method by manually coding an algorithm to identify the appropriate breakpoint. The following code is conceptually very similar to the R version of REGRANS, which was initially written in 1993 in the programming language BASIC [@pezzuto1993].

You can optionally define lower and upper limits for the changepoints you want this function to test; otherwise, it tests a user-defined number of evenly-spaced values (default = 100) ranging from the 0.2 quantile to the 0.8 quantile of the x-axis variable.

```{r}
regrans_est <- regrans(fc, "x", "y", verbose = FALSE)
regrans_est
```

```{r}
#| echo: false

ggplot() +
  geom_point(data = fc, aes(x, y), alpha = 0.4) +
  geom_vline(aes(xintercept = regrans_est, linetype = "Estimate"), linewidth = 1) +
  geom_vline(aes(xintercept = 75, linetype = "Truth"), linewidth = 1) +
  labs(
    title = "REGRANS SM50 Estimate",
    x = "CW (mm)",
    y = "CH (mm)",
    linetype = NULL
  ) +
  mytheme
```

As with the previous broken-stick methods, REGRANS tends to underestimate SM50 and the model estimates are highly sensitive to the values of the upper and lower bounds of the region considered plausible to contain the SM50 value.

## Crab_Maturity Type A (Broken-stick Stevens)

A third algorithm for using segmented regression to estimate SM50 was written by Dr. Bradley Stevens at the University of Maryland Eastern Shore. This code is part of the [Crab_Maturity program](https://github.com/Crabman52/Crustacean_Sensation) available on GitHub. There are several different methods included within Crab_Maturity; this segmented regression approach is included here because it differs from the `segmented` and REGRANS methods in that *possible SM50 values are restricted to values of the x-variable present in the data set*.

```{r}
stevens_est <- broken_stick_stevens(fc, "x", "y", verbose = FALSE)
stevens_est
```

```{r}
#| echo: false
ggplot() +
  geom_point(data = fc, aes(x, y), alpha = 0.4) +
  geom_vline(aes(xintercept = stevens_est, linetype = "Estimate"), linewidth = 1) +
  geom_vline(aes(xintercept = 75, linetype = "Truth"), linewidth = 1) +
  labs(
    title = "Stevens SM50 Estimate",
    x = "CW (mm)",
    y = "CH (mm)",
    linetype = NULL
  ) +
  mytheme
```


### References
