---
title: "Classification methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(morphmat)
```

Here I will write a vignette explaining all of the methods that first classify points as mature or immature, enabling the use of logistic regression to find SM50 as would be done with a physiological metric of maturity. This vignette will be useful for referring readers to more complex options available in packages like `mclust` that are not available within `morphmat`.

# Clustering methods

hclust, kmeans, hkmeans, PAM, OPTICS, DBSCAN, HDBSCAN, Somerton method, mclust

# Other classification methods

## Inflection point discriminant method

# Discriminant Analysis

Talk about drawbacks and reasons why it is unnecessary, but mention instances where it might be useful (stock assessment, applications to related species)

# Post-classification logistic regression

- Basic logistic regression, GLMMs, GAMs, sdmTMB, etc.
- Methods of obtaining standard errors/CIs for SM50 parameter (Delta method, Fieller method, etc.)
- BAYESIAN
