---
title: "morphmat"
output: rmarkdown::html_vignette
link-citations: true
link-external-newwindow: true 
vignette: >
  %\VignetteIndexEntry{morphmat}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
#  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| message: FALSE
#| echo: FALSE
#| warning: FALSE
#| output: FALSE

library(morphmat)
library(dplyr) # tidy data manipulation
library(broom) # cleaning model output
library(ggplot2) # for plotting
library(mclust) # recommended clustering method
```

We will start by simulating a data set with a known SM50 value of 75 mm to demonstrate an example workflow using `morphmat`. See `vignette("simulations")` and the `fake_crustaceans()` documentation for more details on the structure of our artificial data set.
```{r}
#| label: generate-crabs

set.seed(12) # set seed for reproducibility

fc <- fake_crustaceans(
  error_scale = 17,
  slope = 9,
  L50 = 75, # known size at maturity is 75 mm
  n = 800, # sample size
  allo_params = c(0.9, 0.25, 1.05, 0.2),
  x_mean = 85 # mean carapace width of the sample
)

mytheme <- theme_classic() + # define custom theme for ggplots
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
    text = element_text(size = 13))
```

# Plot your data

## Original scale
```{r}
ggplot() +
  geom_point(data = fc, aes(x, y), alpha = 0.4) +
  labs(x = "Carapace width (mm)", y = "Claw height (mm)", ) +
  mytheme
```

## Log-log scale
```{r}
ggplot() +
  geom_point(data = fc, aes(log_x, log_y), alpha = 0.4) +
  labs(x = "Log carapace width (mm)", y = "Log claw height (mm)", ) +
  mytheme
```

# Suggested method

## mclust for classification

```{r}
#| messages: false

clust_dat <- cbind(fc$x, fc$y)
mclust_class <- mclust::Mclust(data = clust_dat,
                               G = 2,
                               modelNames = "EVV")

broom::glance(mclust_class) %>% select(-hypvol)

fc_mclust <- fc %>%
  rename(true_maturity = mature) %>%
  mutate(pred_maturity = mclust_class$classification) 

mature_label <- slice_max(fc_mclust, x) %>% pull(pred_maturity)

fc_mclust <- fc_mclust %>%
  mutate(
    pred_maturity = if_else(pred_maturity == mature_label,
                            as.factor(1), as.factor(0)),
    pred_maturity_num = if_else(pred_maturity == 1,
                                as.numeric(1), as.numeric(0))
  ) %>%
  mutate(uncertainty = mclust_class$uncertainty)
```

## Visualize the mature and immature groups

<!-- # ```{r} -->
<!-- # fc_mclust %>% -->
<!-- #   mutate(mature = if_else(pred_maturity == 0, "Immature", "Mature")) %>% -->
<!-- #   tbl_summary(include = c(x, y), -->
<!-- #     by = mature, -->
<!-- #     type = all_continuous() ~ "continuous2", -->
<!-- #     label = list(x ~ "Carapace width (mm)", y ~ "Chela height (mm)"), -->
<!-- #     statistic = list( -->
<!-- #       all_continuous() ~ c("{mean} (\U00B1{sd})", -->
<!-- #         "{min}-{max}"))) %>% -->
<!-- #   add_p() %>% -->
<!-- #   modify_header(label ~ "", -->
<!-- #     all_stat_cols() ~ "**{level}**<br>N = {n} ({style_percent(p)}%)") %>% -->
<!-- #   add_stat_label(label = all_continuous() ~ c("Mean (\U00B1SD)", "Min - Max")) -->
<!-- # ``` -->

```{r}
#| label: fig-cluster1
#| fig-cap: "Predicted clusters using mixture-based clustering."

ggplot(data = fc_mclust) +
  mytheme +
  geom_point(aes(x, y, color = pred_maturity, fill = pred_maturity),
             alpha = 0.5, shape = 21) +
  scale_color_manual(values = c("0" = "#368aab", "1" = "#993843"),
                     breaks = c(0, 1), labels = c("0" = "Immature", "1" = "Mature")) +
  scale_fill_manual(values = c("0" = "#7bbcd5", "1" = "#CA6E78"),
                    breaks = c(0, 1), labels = c("0" = "Immature", "1" = "Mature")) +
  labs(x = "Carapace width (mm)", y = "Chela height (mm)",
       color = "Predicted clusters", fill = "Predicted clusters")
```

## GLM for obtaining SM50 value

This is the most basic way to obtain an SM50 estimate from allometric data that has been clustered into mature and immature groups:
```{r}
mod <- glm(data = fc_mclust,
           pred_maturity_num ~ x,
           family = binomial(link = "logit"))

unname(-coef(mod)[1] / coef(mod)[2]) # SM50 = -A/B
```

# Comparing clustering approaches
```{r}
all_clusters <- cluster_mods(fc, xvar = "x", yvar = "y", method = c("all"), plot = TRUE)
```

# Comparison with piecewise regression models

```{r}
piecewise_mods(fc, xvar = "x", yvar = "y", method = c("all"))

piecewise_mods(fc, xvar = "x", yvar = "y", method = c("all"),
                   log = TRUE) %>% exp()
```

## Broken-stick/piecewise regression methods - individual functions

REGRANS:

```{r}
regrans(fc, "x", "y", verbose = FALSE)
```

Two-line logistic:

```{r}
two_line_logistic(fc, xvar = "x", yvar = "y", verbose = FALSE, SM50_start = 85)
```

Two-line model (lines are fit separately; no forced intersection):

```{r}
two_line(fc, xvar = "x", yvar = "y", verbose = FALSE)
```

Broken-stick Stevens (only iterates over values of the x-axis variable present in the data):

```{r}
broken_stick_stevens(fc, xvar = "x", yvar = "y", verbose = FALSE)
```
