---
title: "Post-classification logistic regression"
link-citations: true
link-external-newwindow: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
 # collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}

# List of packages required:
packages <- c("morphmat", "broom", "drc", "drda", "qra", "lmtest", "sandwich",
  "mclust", "dplyr", "ggplot2")

# Load packages into session
lapply(packages, require, character.only = TRUE)

mytheme <- theme_classic() + # define custom theme for ggplots
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
    text = element_text(size = 13))
```

```{r}
set.seed(12) # set seed for reproducibility

fc <- fake_crustaceans(
  error_scale = 17,
  slope = 9,
  L50 = 75, # known size at maturity is 75 mm
  n = 800, # sample size
  allo_params = c(0.9, 0.25, 1.05, 0.2),
  x_mean = 85 # mean carapace width of the sample
)

clust_dat <- cbind(fc$x, fc$y)
mclust_class <- mclust::Mclust(data = clust_dat,
                               G = 2,
                               modelNames = "EVV")

fc_mclust <- fc %>%
  rename(true_maturity = mature) %>%
  mutate(pred_maturity = mclust_class$classification) 

mature_label <- slice_max(fc_mclust, x) %>% pull(pred_maturity)

fc_mclust <- fc_mclust %>%
  mutate(
    pred_maturity = if_else(pred_maturity == mature_label,
                            as.factor(1),
                            as.factor(0)),
    pred_maturity_num = if_else(pred_maturity == 1,
                                as.numeric(1),
                                as.numeric(0))) %>%
  mutate(uncertainty = mclust_class$uncertainty)
```

# Post-classification logistic regression

Beyond the familiar `stats::glm`, there are many packages that can be used to model the relationship between size and maturity status, allowing for the incorporation of random effects, temporal or spatial structuring, and other additional complexities. These include popular packages for fitting generalized linear mixed models (GLMMs) and generalized additive (mixed) models (GAMMs) such as `lmer`, `nlme`, `mgcv`, `glmmTMB`, and `sdmTMB`. While logit links are the most common for binomial models like these, testing alternative link functions—particularly the probit and complementary log-log ("cloglog") links—may be important for ensuring reliable parameter estimates [@mainguy2024]. Also see Roa et al. [-@roa1999].

Helpful vignette: <https://cran.r-project.org/web/packages/qra/vignettes/timeMortality.html>

Even if no covariates are included in the model, many choices must be made when fitting a binomial model for the probability of maturity at a given length. These include:

- The model to be fitted (most commonly 2-parameter logistic models, but many alternative formulations with up to 5 parameters)
- Alternative link functions (logit, probit, cloglog, etc.)
- Constrained optimization (i.e., set lower and upper bounds on the possible values of the parameters to be estimated)
- Initial values provided for the parameters to be estimated


## Example methods to obtain confidence intervals for SM50 value

A great place to start is the [confint_L.R script](https://github.com/rafamoral/L50/blob/main/confint_L.R) from Mainguy et al. [-@mainguy2024]. This script can be used to quantify the uncertainty around the SM50 parameter using the Delta method, Fieller method, profile-likelihood, three resampling techniques (non-parametric bootstrapping, parametric bootstrapping, and Monte Carlo), and Bayesian credible intervals. For the resampling methods, three types of intervals can be calculated: equal-tailed intervals (ETI/percentile), bias corrected accelerated intervals (BCa), or highest density intervals (HDI).

### Standard glm

```{r}
example_glm <- glm(data = fc_mclust,
                   pred_maturity_num ~ x,
                   family = binomial(link = "logit"))

broom::tidy(example_glm, conf.int = TRUE)

broom::glance(example_glm) %>%
  dplyr::select(-c(null.deviance, df.null)) %>%
  mutate(across(where(is.double), ~ round(.x, digits = 2)))

```

### drc package

The R package [drc](https://cran.r-project.org/web/packages/drc/drc.pdf) was designed for analysis of dose-response data, and includes many functions that are useful for fitting logistic models [@ritz2015]. This package can fit a variety of 5/4/3/2-parameter logistic, log-logistic, and Weibull models and find confidence intervals for the estimated parameters, among other functions.

```{r}
testdrm <- drc::drm(pred_maturity_num ~ x,
               data = fc_mclust,
               fct = LL.2(),
               type = "binomial")
testdrm
summary(testdrm)
confint(testdrm, "e") # specifies the 'e' parameter, the inflection point

estfun.drc <- drc::estfun.drc
bread.drc <- drc::bread.drc

# use lmtest and sandwich packages to obtain robust standard errors
lmtest::coeftest(testdrm, vcov = sandwich) 
drc::ED(testdrm, c(50), "delta")

plot(testdrm) # built-in plotting method
```

Custom plot example:

```{r}
sm50_est <- coef(testdrm)[2]
low_ci <- confint(testdrm, "e")[1]
hi_ci <- confint(testdrm, "e")[2]

ggplot() +
  geom_point(aes(x = fc_mclust$x, y = fc_mclust$pred_maturity_num),
             shape = 1, size = 1) +
  geom_line(aes(x = fc_mclust$x, y = predict(testdrm))) +
  geom_vline(xintercept = sm50_est) +
  geom_rect(aes(xmin = low_ci, xmax = hi_ci,
    ymin = -Inf, ymax = Inf, fill = "95% CI"), alpha = 0.5) +
  labs(y = "Probability of maturity", x = "Carapace width (mm)", fill = NULL) +
  mytheme +
  theme(legend.position = "inside", legend.position.inside = c(0.75, 0.25))
```

### DRDA package

The R package DRDA is a newer alternative to drc that uses an improved optimization methodology to improve the accuracy of parameter estimates [@malyutina2023].

```{r}
fit_l2 <- drda(pred_maturity_num ~ x,
               data = fc_mclust,
               mean_function = "logistic2")
fit_l4 <- drda(pred_maturity_num ~ x,
               data = fc_mclust,
               mean_function = "logistic4")

anova(fit_l2, fit_l4)

summary(fit_l2)

plot(fit_l2) # built-in plotting method
```

Custom plot example:

```{r}
sum_l2 <- summary(fit_l2)[["param"]][4,]

sm50_est <- sum_l2[1]
low_ci <- sum_l2[3]
hi_ci <- sum_l2[4]

ggplot() +
  geom_point(aes(x = fc_mclust$x, y = fc_mclust$pred_maturity_num),
             shape = 1, size = 1) +
  geom_line(aes(x = fc_mclust$x, y = predict(fit_l2))) +
  geom_vline(xintercept = sm50_est) +
  geom_rect(aes(xmin = low_ci, xmax = hi_ci,
    ymin = -Inf, ymax = Inf, fill = "95% CI"), alpha = 0.5) +
  labs(y = "Probability of maturity", x = "Carapace width (mm)", fill = NULL) +
  mytheme +
  theme(legend.position = "inside", legend.position.inside = c(0.75, 0.25))
```

### qra package

The R package [qra](https://cran.r-project.org/web/packages/qra/index.html) is another useful package initially designed for analyzing dose-response data [@maindonald2021]. It includes implementations of the the Fieller and Delta methods for estimating confidence intervals for the SM50 parameter when it is not directly estimated during model-fitting. See `qra::fieller()` for details of usage and outputs.

```{r}
qra::fieller(
  0.5,
  b = coef(example_glm),
  vv = vcov(example_glm),
  type = "Fieller"
)

qra::fieller(
  0.5,
  b = coef(example_glm),
  vv = vcov(example_glm),
  type = "Delta" # probably OK because of low g value
)
```

Custom plot example:

```{r}
qra_est <- qra::fieller(
  0.5,
  b = coef(example_glm),
  vv = vcov(example_glm),
  type = "Fieller"
)

sm50_est <- qra_est[1]
low_ci <- qra_est[3]
hi_ci <- qra_est[4]

ggplot() +
  geom_point(aes(x = fc_mclust$x, y = fc_mclust$pred_maturity_num),
             shape = 1, size = 1) +
  geom_line(aes(x = fc_mclust$x, y = predict(example_glm, type = "response"))) +
  geom_vline(xintercept = sm50_est) +
  geom_rect(aes(xmin = low_ci, xmax = hi_ci,
    ymin = -Inf, ymax = Inf, fill = "95% CI"), alpha = 0.5) +
  labs(y = "Probability of maturity", x = "Carapace width (mm)", fill = NULL) +
  mytheme +
  theme(legend.position = "inside", legend.position.inside = c(0.75, 0.25))
```

### Others

Other functions that could be used (that I have not tested) include [`twopartm::FiellerRatio()`](https://cran.r-project.org/web/packages/twopartm/index.html) and [`tidydelta::tidydelta()`](https://javiermtzrdz.github.io/tidydelta/).


## References
